\chapter{Sheaf-Theoretic Foundations of Semantic Repair}
\label{ch:sheaf}

\epigraph{The Yarncrawler is a sheaf-based repair machine. Local parsing windows
correspond to stalks, repair corresponds to introducing new morphisms, and entropy
corresponds to cohomological obstruction.}{}

\section{Semantic Space and Covers}

Let $X$ denote a semantic space---the domain of cultural landmarks, neural states,
semantic nodes, or any other structured collection of meaning-bearing objects. A
cover $\mathcal{U} = \{U_i\}_{i \in I}$ represents overlapping neighborhoods of
$X$, each corresponding to a local parsing window within which semantic content
can be coherently described.

\subsection*{Toy Example: Three Overlapping Interpretive Windows}

Before the general definition, consider the simplest case that exhibits a gluing
failure. Let $X = [0,3]$ represent a temporal sequence of observations, covered
by three overlapping windows:
\[
  U_1 = [0, 1.5], \quad U_2 = [1, 2.5], \quad U_3 = [2, 3].
\]
Each window $U_i$ contains a local section $s_i$: a semantic interpretation of
what was observed during that interval. Suppose:
\begin{itemize}
  \item $s_1$ concludes: the object encountered is a seed.
  \item $s_2$ concludes: the object encountered is a pebble.
  \item $s_3$ concludes: the object is food.
\end{itemize}
On the overlap $U_1 \cap U_2 = [1, 1.5]$, both $s_1$ and $s_2$ must agree about
what was happening during that interval. But ``seed'' and ``pebble'' are
incompatible. This incompatibility is a \emph{Čech 1-cocycle}: a mismatch on
an overlap that prevents gluing into a consistent global interpretation.

A repair operation $\rho_{12}$ introduces the softened category ``seed-like
object'' into $s_2$, which is compatible with both. The cocycle is trivialized:
the 1-cocycle becomes a coboundary. On the overlap $U_2 \cap U_3 = [2, 2.5]$,
``seed-like object'' and ``food'' are compatible under the coarser category
``edible item.'' A global section $s$ on all of $[0,3]$ now exists: the
observation was of an edible, seed-like item.

Crucially, the system defers collapse. Until $U_3$ is observed, multiple global
sections remain viable (``pebble'' vs.\ ``seed-like''). The non-uniqueness of
the global section is captured by the non-triviality of $H^1$. Only when $U_3$
provides the additional constraint ``food'' does the ambiguity resolve. Strategic
ambiguity---holding multiple viable global sections in suspension---is the correct
epistemic response to underdetermined gluing.

\begin{definition}{Presheaf of Semantic Modules}
A \emph{presheaf of semantic modules} is a contravariant functor
\[
  \mathcal{S}: \mathcal{U}^{\rm op} \to \mathbf{Cat}
\]
that assigns to each open neighborhood $U \subseteq X$ a category $\mathcal{S}(U)$
of local semantic modules. Objects of $\mathcal{S}(U)$ are local semantic threads,
claims, cached structures, or repaired routines valid within $U$. Morphisms are
semantic restrictions or rewrites. For inclusions $V \subseteq U$, the restriction
functor $\rho^U_V: \mathcal{S}(U) \to \mathcal{S}(V)$ encodes context reduction:
what is asserted in the larger context may be forgotten or specialized in the
smaller one.
\end{definition}

\section{The Sheaf Condition and Semantic Repair}

\begin{definition}{Semantic Tear and Repair}
A family of local sections $\{s_i \in \mathcal{S}(U_i)\}$ is \emph{glueable} if:
(i) local consistency holds: $\rho^{U_i}_{U_i \cap U_j}(s_i) \cong
\rho^{U_j}_{U_i \cap U_j}(s_j)$ for all overlapping pairs; and (ii) there exists
a global extension $s \in \mathcal{S}(\bigcup_i U_i)$ restricting to each $s_i$.

A \emph{semantic tear} occurs when local consistency fails on some overlap
$U_i \cap U_j$: the restriction of $s_i$ and the restriction of $s_j$ to their
common domain are not isomorphic. A \emph{repair} is the introduction of new objects
or morphisms into $\mathcal{S}(U_i)$ or $\mathcal{S}(U_j)$ such that glueability
is restored.
\end{definition}

\section{Cohomology as Semantic Entropy}

Cohomology detects the unresolved incompatibilities that remain after local
repairs have been applied. The Čech cohomology groups of the presheaf
$\mathcal{S}$ with respect to the cover $\mathcal{U}$ organize these
obstructions by dimension:
\begin{align*}
  H^0(X, \mathcal{S}) &\quad \text{viable global sections (coherent meanings)}, \\
  H^1(X, \mathcal{S}) &\quad \text{minimal ambiguities (semantic seams)}, \\
  H^k(X, \mathcal{S}),\; k \geq 2 &\quad \text{deeper obstructions to coherence.}
\end{align*}

These cohomology groups are interpreted as measures of semantic entropy: the
unresolved cocycles represent ambiguity stored for potential reinterpretation. Repair
corresponds to introducing new morphisms that trivialize cocycles, moving from
$H^1$ toward $H^0$ by resolving the incompatibilities that the cocycles record.

\begin{remark}
Strategic ambiguity is a resource rather than a failure. A Yarncrawler that
collapses all ambiguity immediately---forcing a unique global section at every
step---forfeits the adaptability that deferred gluing provides. The presence of
a non-trivial $H^1$ is a signal that multiple viable interpretations exist; the
system defers collapse until additional evidence arrives to determine which
interpretation to commit to.
\end{remark}

\section{The Stigmergic Repair Closure Theorem}

\begin{theorem}{Stigmergic Repair Closure}
Let $\mathcal{S}$ be a presheaf of local semantic sections over patches
$\{U_i\}$, with overlaps $U_i \cap U_j$ that may admit inconsistent
interpretations. Suppose that for each overlap there exists an elastic repair
operation $\rho_{ij}$ satisfying:
\begin{enumerate}
  \item $\rho_{ij}$ preserves type safety: objects remain within a coarser
        semantic category (for example, ``seed-like'' rather than either
        ``seed'' or ``pebble'');
  \item $\rho_{ij}$ reduces local free energy $\mathcal{F}(U_i \cap U_j)$
        relative to the unrepaired sections; and
  \item $\rho_{ij}$ is stigmergic: the cost of subsequent repairs decreases
        with the density of prior contributions.
\end{enumerate}
Then the family of repaired sections $\{s^{\rho}_i\}$ admits at least one global
section $s$ with $\mathcal{F}(s) \leq \sum_i \mathcal{F}(s_i)$ and with viability
increasing in the cumulative density of repairs.
\end{theorem}

\begin{proposition}{Repair Preserves Blanket Structure}
If every repair is generated by local morphisms that commute with restrictions and
respect conditional independence at the Markov blanket---that is, if
$X_t \perp\!\!\!\perp E_t \mid (S_t, A_t)$ is maintained throughout---then any
finite composition of repairs preserves the gluing structure of $\mathcal{S}$.
\end{proposition}

\section{Self-Maintaining Growth under Stigmergic Feedback}

\begin{theorem}{Self-Maintaining Growth}
Let $\Phi(t)$ denote the accumulated density of a repaired structure (cache, berm,
semantic section). Suppose the repair cost function $C(n)$ for the $n$-th
contribution is monotonically decreasing in cumulative density $\Phi(n)$. Then
there exists a critical threshold $\Phi^*$ such that
\[
  \Phi(n+1) - \Phi(n) \geq \delta > 0 \quad \text{whenever } \Phi(n) \geq \Phi^*,
\]
that is, once $\Phi^*$ is surpassed, the system enters a regime of self-sustaining
growth where each new repair lowers the barrier for future repairs.
\end{theorem}

The discrete-time dynamics implementing this theorem take the form:
\[
  \Phi_{t+1} = \Phi_t + \Delta t \left(
    \alpha \frac{\Phi_t}{\Phi_t + \theta}
    \left(1 - \frac{\Phi_t}{K}\right) - \delta \Phi_t
  \right),
\]
where $\alpha \equiv \lambda \kappa$ is baseline inflow, $\theta > 0$ encodes early
friction (the cost of initial contributions), $K$ is carrying capacity, and
$\delta \geq 0$ is decay. The stigmergic threshold condition is $\alpha > \delta\theta$.

\section{Ecological and Cultural Instantiations}

The abstract sheaf-repair mechanism is grounded by two concrete instantiations that
run across the ecological and cultural domains analyzed in the companion work.

In squirrel foraging, each foraging episode is a local patch $U_i$ with a section
$s_i$ encoding the interpretation of an encountered object. A nut recognized in
one context and a visually similar pebble encountered in another produce conflicting
restrictions at the overlap. Rather than collapsing, the squirrel repairs by adopting
a softened category---``seed-like object.'' The cocycle condition fails strictly but
is restored under a relaxed gluing criterion, preserving viability. Multiple possible
global interpretations are held in suspension until additional evidence resolves the
ambiguity.

In berm construction, an analogous process occurs at the cultural scale. A villager
encountering a cairn at a territorial edge may perceive it as insufficient and add
more material. The material composition need not be uniform; bones, ash, and dung
may be included alongside rocks. Local repairs introduce semantic elasticity: the
criterion is ``boundary-marker-like'' rather than ``stone-only.'' Over time,
overlapping deposits cohere into self-maintaining structures that attract further
contributions, forming the Voronoi-like tessellations of Amazonian terra preta.

In both cases, the global section---the food cache or the territorial berm---emerges
not from top-down planning but from the gluing of locally inconsistent but compatible
acts of repair. This is the universal mechanism of the Yarncrawler framework: local
disagreements reconciled through semantic elasticity, strategic ambiguity, and
stigmergic reinforcement.
