\chapter*{Prefatory Note: On the Form of This Text}
\addcontentsline{toc}{chapter}{Prefatory Note}

This monograph does not proceed from a completed theoretical edifice toward illustrative
applications. It proceeds in the opposite direction: from distributed conceptual material toward
progressive topological stabilization. The reader should expect definitions to accumulate rather
than appear fully formed, and should treat apparent repetition across sections as triangulation
rather than redundancy.

The writing style is a mechanistic consequence of the theory rather than a stylistic preference.
The core argument of Part~\ref{part:method} is that any text arguing for provenance preservation
and reconstruction transparency while itself operating through aggressive compression and assumed
priors would be internally inconsistent. The prefatory note, the proliferating appendices, the
deliberate restating of assumptions at each new level of abstraction are not ornamental scholarly
rituals. They are consequences of taking seriously the claim that meaning is reconstructed rather
than transmitted.

Most of the conceptual work assembled here already existed as a large-scale distributed semantic
field before this document imposed topological structure on it. Different companion papers
foreground different regions of the same underlying geometry: cognition, cosmology, social
systems, abstraction, entropy, compression, trajectory selection, admissibility, reconstruction,
provenance. The apparent diversity of subjects conceals a relatively small set of deep structural
operations being reapplied across domains and scales.

A note on formal notation. Definitions, theorems, propositions, and axioms are presented in
numbered environments so that dependency relations can be tracked explicitly. Where a later
argument depends on an earlier definition, the dependency is marked. Where a concept is
imported from a companion document rather than derived here, the source is indicated.
Terminology is introduced progressively and is not assumed stable across disciplinary
boundaries.

\vspace{1cm}
\begin{flushright}
\itshape Flyxion\\
Canada, 2026
\end{flushright}
\cleardoublepage
