% ============================================================
%  THE ORTYX PROTOCOL
%  main.tex — master compilation file
%
%  Engine: LuaLaTeX (TeX Live 2023)
%  Run twice for cross-references and bookmarks.
%
%  Build command:
%    lualatex -interaction=nonstopmode main.tex
% ============================================================

\documentclass[12pt,twoside,openright]{book}

% ── Preamble modules (order matters) ────────────────────────
\input{preamble/packages}
\input{preamble/colors}
\input{preamble/typography}
\input{preamble/theorem-envs}
\input{preamble/boxes}
\input{preamble/macros}

% ============================================================
\begin{document}
% ============================================================

% ── Front matter ────────────────────────────────────────────
\frontmatter

% Title page
\begin{titlepage}
\thispagestyle{empty}
\vspace*{0.10\textheight}
{\centering
  {\color{Ortyx@subtle}\rule{0.55\textwidth}{0.4pt}}\par
  \vspace{20pt}
  {\color{Ortyx@rule}\fontsize{28}{34}\selectfont\bfseries
    The Ortyx Protocol}\par
  \vspace{16pt}
  {\color{Ortyx@neutral}\large\itshape
    Decomposition, Preservation, and\\[4pt]
    Constraint-Preserving Reinterpretation\\[4pt]
    of Speculative Computational Systems}\par
  \vspace{22pt}
  {\color{Ortyx@subtle}\rule{0.55\textwidth}{0.4pt}}\par
  \vspace{36pt}
  {\color{Ortyx@neutral}\Large Flyxion}\par
  \vspace{12pt}
  {\color{Ortyx@neutral}\normalsize 2026}\par
  \vfill
  {\color{Ortyx@neutral}\small\itshape
    Independent Research}\par
  \vspace{20pt}
}
\end{titlepage}

% Copyright / imprint page
\thispagestyle{empty}
\vspace*{\fill}
{\small\color{Ortyx@neutral}
\noindent
\textit{The Ortyx Protocol:\\
Decomposition, Preservation, and Constraint-Preserving Reinterpretation\\
of Speculative Computational Systems}

\medskip
\noindent
Copyright \textcopyright{} 2026 Flyxion. All rights reserved.

\medskip
\noindent
Published independently under the name Flyxion.\\
GitHub: \texttt{standardgalactic}

\medskip
\noindent
This work may be freely shared for non-commercial scholarly purposes
with attribution. No part may be reproduced for commercial purposes
without written permission.

\medskip
\noindent
\textit{First edition, 2026.}

\vspace{16pt}
\noindent
\textit{Note on the title:} The ortyx is a bird related to the quail,
known in ancient sources for its migratory discipline and capacity to
correct against drift without external landmarks. The name is used here
as an image of directed movement through uncertain epistemic terrain ---
not by the confidence of a system that knows its destination, but by the
discipline of one that knows when it is drifting.
}
\vspace*{\fill}
\clearpage

% Epigraph page
\thispagestyle{empty}
\vspace*{0.28\textheight}
{\centering
  \begin{minipage}{0.72\textwidth}
    \centering
    \color{Ortyx@neutral}
    \itshape
    ``The test of a framework is not what it can explain,\\
    but what it would forbid. A theory that forbids nothing\\
    is not a theory; it is an atmosphere.''

    \vspace{12pt}
    {\small\upshape --- Flyxion}
  \end{minipage}
}
\vspace*{\fill}
\clearpage

% Table of contents
\tableofcontents

% Preface
\include{frontmatter/preface}

% ── Main matter ─────────────────────────────────────────────
\mainmatter

% ════════════════════════════════════════════════════════════
%  PART I — IMMERSION
% ════════════════════════════════════════════════════════════
\partinterlude{Part I}{Immersion}{%
The Phoenix ecosystem is encountered on its own terms.
Reconstruction precedes judgment. The framework is allowed to
demonstrate its explanatory power, its compression advantages,
and its architectural coherence before any pressure is applied.
The reader who finds the system compelling at this stage is
responding correctly. That response is the starting point,
not an error to be corrected.}

\include{parts/part1/ch01}
\include{parts/part1/ch02}
\include{parts/part1/ch03}
\include{parts/part1/ch04}
\include{parts/part1/ch05}

% ════════════════════════════════════════════════════════════
%  PART II — TENSION
% ════════════════════════════════════════════════════════════
\partinterlude{Part II}{Tension}{%
The audit voice is born without being named. Asymmetries accumulate:
some equations constrain outcomes, others merely accompany prose;
some concepts compress phenomena, others absorb them
indiscriminately. The reader should feel structural stress before
the vocabulary for naming it becomes available.}

\include{parts/part2/ch06}
\include{parts/part2/ch07}
\include{parts/part2/ch08}
\include{parts/part2/ch09}
\include{parts/part2/ch10}

% ════════════════════════════════════════════════════════════
%  PART III — DECOMPOSITION
% ════════════════════════════════════════════════════════════
\partinterlude{Part III}{Decomposition}{%
The audit operators are deployed. Four failure geometries are
defined and illustrated. The four inferential levels are
disentangled. This is the most formally demanding section of
the book. RSVP is also partially audited here. Nothing passes
through decomposition unexamined.}

\include{parts/part3/ch11}
\include{parts/part3/ch12}
\include{parts/part3/ch13}
\include{parts/part3/ch14}
\include{parts/part3/ch15}

% ════════════════════════════════════════════════════════════
%  INTERLUDE — WHAT SURVIVES COLLAPSE
% ════════════════════════════════════════════════════════════
\include{parts/interlude-collapse/interlude}

% ════════════════════════════════════════════════════════════
%  PART IV — RECONSTRUCTION
% ════════════════════════════════════════════════════════════
\partinterlude{Part IV}{Reconstruction}{%
The reinterpretation functor $\RSVPfunctor : \ThetaS \to \ThetaStar$
is applied to the survivable components. RSVP does not win by
assertion; it wins because it survives the audit with fewer
projection defects and lower recursive closure metrics. The
question is not whether RSVP is true, but whether the surviving
structures become more tractable in its language.}

\include{parts/part4/ch16}
\include{parts/part4/ch17}
\include{parts/part4/ch18}
\include{parts/part4/ch19}
\include{parts/part4/ch20}

% ════════════════════════════════════════════════════════════
%  PART V — METHODOLOGICAL SYNTHESIS
% ════════════════════════════════════════════════════════════
\partinterlude{Part V}{Methodological Synthesis}{%
The Ortyx Protocol is stated explicitly and subjected to the same
recursive pressure it applies to everything else.
$\OrtOp_{n+1} = \AuditOp(\OrtOp_n)$. The book ends not with a
stable doctrine but with a protocol that survives only through
continued self-application. That recursion is not a weakness.
It is the only honest form a methodology of this kind can take.}

\include{parts/part5/ch21}
\include{parts/part5/ch22}
\include{parts/part5/ch23}
\include{parts/part5/ch24}

% ── Appendices ──────────────────────────────────────────────
\appendix

\include{appendices/appA}
\include{appendices/appB}
\include{appendices/appC}
\include{appendices/appD}
\include{appendices/appE}
\include{appendices/appF}
\include{appendices/appG}

% ── Back matter ─────────────────────────────────────────────
\backmatter

% Bibliography
\bibliographystyle{plainnat}
\bibliography{bibliography/ortyx}

\end{document}
