% ============================================================
%  THE ORTYX PROTOCOL
%  Appendix C: Harmonic Consensus and Sparse Relational Compression
% ============================================================

\chapter{Harmonic Consensus and Sparse Relational Compression}
\label{app:harmonic}

\noindent
This appendix develops the formal mathematics of the
HCF1 harmonic consensus compression framework. The
encoding efficiency claims that survive the decomposition
at \LevelI{} are derived here.

\section{Independent Harmonic Encoding}
\label{app-c:independent}

Let a signal contain a harmonic complex with fundamental
frequency $f_0$ and $N$ harmonics at frequencies
$f_n = n f_0$, $n = 1, \ldots, N$. Let $A_n(t)$ and
$\phi_n(t)$ denote the amplitude envelope and instantaneous
phase of the $n$-th harmonic respectively. The
traditional spectral encoding stores all harmonics
independently:
\[
  \mathcal{H} = \{(A_n(t), f_n, \phi_n(t))\}_{n=1}^{N}.
\]
The storage complexity is $O(N)$ per time step.

\section{The Conformity Relation}
\label{app-c:conformity}

Define the normalized amplitude velocity of the $n$-th
harmonic:
\[
  v_n(t) = \frac{dA_n/dt}{A_n(t)}.
\]
Define the harmonic conformity relation with conformity
threshold $\varepsilon > 0$:
\[
  \Gamma_n(t) = \mathbf{1}\!\left(|v_n(t) - v_0(t)| < \varepsilon\right).
\]
When $\Gamma_n(t) = 1$, harmonic $n$ is evolving in
proportion to the fundamental within tolerance
$\varepsilon$. When $\Gamma_n(t) = 0$, harmonic $n$
is deviating from the fundamental's envelope dynamics.

\begin{definition}
A harmonic complex is \emph{$\varepsilon$-conformant}
at time $t$ if $\sum_{n=1}^{N} \Gamma_n(t) = N$ ---
that is, all harmonics conform to the fundamental
within tolerance $\varepsilon$.
\end{definition}

\section{Sparse Consensus Encoding}
\label{app-c:consensus}

The sparse consensus encoding stores:
\[
  \mathcal{H}_{\mathrm{cons}}
  = \{A_0(t), f_0, \phi_0(t), \Gamma_1(t), \ldots, \Gamma_N(t)\}
  \cup \{(A_n(t), \delta\phi_n(t)) : \Gamma_n(t) = 0\},
\]
where $\delta\phi_n(t)$ is the phase deviation of
non-conforming harmonics from prediction. Let $k(t)$
denote the number of non-conforming harmonics at time
$t$: $k(t) = N - \sum_{n=1}^N \Gamma_n(t)$.

\begin{theorem}
\label{thm:compression}
The consensus encoding has storage complexity $O(1) + O(k(t))$
per time step, compared to $O(N)$ for independent
encoding. The compression ratio is:
\[
  \rho(t) = \frac{O(N)}{O(1) + O(k(t))} \approx \frac{N}{k(t)}
\]
for large $N$ and small $k(t)$. For strongly coupled
harmonic sources with high conformity rates $k(t)/N \ll 1$,
the compression ratio is large.
\end{theorem}

\section{Expected Compression Over Time}
\label{app-c:expected}

For a stationary stochastic model where each harmonic
independently conforms with probability $p_{\mathrm{conf}}$:
\[
  \mathbb{E}[k(t)] = N(1 - p_{\mathrm{conf}}).
\]
The expected storage complexity under consensus encoding is:
\[
  \mathbb{E}[\text{cost}] = O(1) + O(N(1 - p_{\mathrm{conf}})).
\]
For strongly coupled harmonic sources (high physical
coupling, $p_{\mathrm{conf}} \to 1$), the expected
cost approaches $O(1)$. For uncoupled sources
($p_{\mathrm{conf}} \to 0$), the expected cost approaches
$O(N)$, identical to independent encoding.

\section{Connection to Predictive Coding}
\label{app-c:predictive-coding}

The consensus encoding can be interpreted as a predictive
code in the following sense. The predictor for harmonic
$n$ at time $t$ is:
\[
  \hat{A}_n(t) = A_n(t - \Delta t) \cdot
  \exp\!\bigl(v_0(t)\,\Delta t\bigr),
\]
the prediction that harmonic $n$ evolves proportionally
to the fundamental over the interval $\Delta t$.
The conformity indicator $\Gamma_n(t)$ tests whether
the prediction error falls within tolerance $\varepsilon$.
The consensus encoding stores only the prediction errors
that exceed the tolerance threshold, in the same way
that predictive coding in the signal processing literature
stores only the residual between predicted and actual values.

\begin{proposition}
The consensus encoding achieves the entropy of the
conformity process $\{\Gamma_n(t)\}$ plus the entropy
of the residuals $\{A_n - \hat{A}_n : \Gamma_n = 0\}$,
which is at most the entropy of the full independent
encoding whenever the conformity rate exceeds $1/2$.
\end{proposition}
