% ============================================================
%  THE ORTYX PROTOCOL
%  Appendix F: Recursive Semantic Drift and
%              Representational Closure
% ============================================================

\chapter{Recursive Semantic Drift and Representational Closure}
\label{app:closure}

\noindent
This appendix formalizes the recursive closure dynamics
identified informally in Chapter~6 and measured by
the $\AuditR$ operator in Chapter~11. The semantic
closure metric provides a quantitative basis for
identifying when a framework's vocabulary is evolving
primarily under its own dynamics rather than under
corrective pressure from the world.

\section{The Representation--Reality Gap}
\label{app-f:gap}

Let $X_t \in X$ denote the underlying system state
at time $t$ and $M_t \in M$ represent the framework's
current reduced symbolic state (its vocabulary, its
current interpretive commitments, its active claims).
Suppose the representation evolves under a self-driven
mapping $F : M \to M$ and an environmental correction
term with coupling strength $\kappa \ge 0$:
\[
  M_{t+1} = F(M_t) + \kappa\,G(X_t),
\]
where $G : X \to M$ maps the underlying state to a
correction signal in the representational space.

\section{The Semantic Closure Metric}
\label{app-f:closure-metric}

\begin{definition}
The \emph{semantic closure metric} at time $t$ is:
\[
  \SemClose(t) = \frac{\|F(M_t)\|}{\|G(X_t)\|},
\]
measuring the ratio of internal dynamics to external
correction. When $\SemClose(t) \gg 1$, the representation
evolves primarily under its own dynamics; when
$\SemClose(t) \ll 1$, the representation is dominated
by external correction.
\end{definition}

\begin{definition}
\emph{Runaway closure} occurs when:
\[
  \SemClose(t) \to \infty \quad \text{as} \quad t \to \infty,
\]
regardless of the underlying system state $X_t$. This
corresponds to the representation becoming fully self-referential:
its evolution is determined by its own dynamics, not
by the phenomena it purports to represent.
\end{definition}

\section{Conditions for Closure}
\label{app-f:conditions}

Runaway closure occurs under the following conditions:

\begin{enumerate}
  \item \textit{Weak coupling} ($\kappa \to 0$): external
  correction becomes negligible, leaving the representation
  to evolve under $F$ alone.
  \item \textit{Strong internal dynamics} ($\|F(M_t)\|$
  grows faster than $\|G(X_t)\|$): the self-driven
  dynamics outpace the corrective pressure from reality.
  \item \textit{Accommodating $F$}: the internal mapping
  $F$ is structured to preserve vocabulary coherence
  regardless of input --- it absorbs any $M_t$ into
  a smooth continuation.
\end{enumerate}

In the context of the \Ortyx{} audit, condition~(3)
corresponds to $\FailGeo{3}$ (Semantic Universality
Collapse): the mapping $F$ has been defined broadly
enough to accommodate any incoming $M_t$ without
generating internal inconsistency, which means external
correction can never produce an inconsistency that
would force revision.

\section{The $\AuditR$ Operator}
\label{app-f:audit-r}

The $\AuditR$ operator is a summary statistic measuring
the degree of closure risk in a framework $\SysTriple$:

\[
  \AuditR(\SysTriple) = \mathbb{E}\bigl[\SemClose(t)\bigr]_T,
\]
where the expectation is taken over a test period $T$
during which the framework is subjected to representative
challenges. In practice, $\AuditR$ is assessed qualitatively
by examining:

\begin{enumerate}
  \item Whether the framework's vocabulary has migrated
  across domains in ways that weaken external traction.
  \item Whether the framework has developed accommodation
  strategies that render potential disconfirmations
  consistent with its claims.
  \item Whether the framework's self-description is
  becoming the primary input to its future development
  rather than empirical evidence.
\end{enumerate}

High $\AuditR$ scores indicate that the framework is
approaching the closure regime; the appropriate response
is to increase $\kappa$ by introducing more demanding
external falsification tests rather than to allow the
internal dynamics to continue unchecked.
