\documentclass[10pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage{mathpazo}

\usepackage{amsmath,amssymb,amsthm,mathtools}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\usepackage{stmaryrd}
\usepackage{tikz}
\usetikzlibrary{decorations.pathmorphing}
\usepackage{tikz-cd}
\usepackage{mathrsfs}
\usepackage{geometry}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multicol}
\usepackage{physics}
\usepackage{bm}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{mdframed}
\usepackage{titlesec}

\geometry{margin=0.65in, top=0.75in, bottom=0.75in}

\pagestyle{empty}

\setlength{\parindent}{0pt}
\setlength{\parskip}{0.3em}
\setlength{\columnsep}{1.2em}
\raggedcolumns

% Section formatting
\titleformat{\section}[block]
  {\normalfont\bfseries\small}
  {}
  {0pt}
  {\rule{\linewidth}{0.4pt}\\\vspace{0.1em}}
  [\vspace{0.1em}]

% Shaded box for compression statements
\mdfdefinestyle{compbox}{
  backgroundcolor=gray!10,
  linecolor=gray!40,
  linewidth=0.5pt,
  innerleftmargin=5pt,
  innerrightmargin=5pt,
  innertopmargin=4pt,
  innerbottommargin=4pt,
  skipabove=4pt,
  skipbelow=4pt
}

% ---- Macros ----
\newcommand{\A}{\mathcal{A}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\fM}{\mathfrak{M}}
\newcommand{\Pp}{\mathcal{P}}
\newcommand{\Rr}{\mathcal{R}}
\newcommand{\Svalid}{\mathcal{S}_{\mathrm{valid}}}
\newcommand{\Adm}{\mathcal{A}_{\mathrm{dm}}}
\newcommand{\Hcech}{\check{H}}
\newcommand{\KES}{\mathfrak{S}}

\begin{document}

\begin{center}
{\LARGE \textbf{The Kinetic--Event Synthesis (KES)}}\\[0.4em]
{\large Mathematical Supplement \& Structural Reference}\\[0.2em]
{\small Flyxion \quad\textbullet\quad Independent Researcher}
\end{center}

\vspace{0.6em}

\begin{multicols}{2}

% ============================================================
\section*{1.\enspace Core KES State}

\[
\KES_t =
(X_t,\,\Omega_t,\,H_t,\,\Psi_t,\,\M_t,\,\G,\,\mathcal{F})
\]

\[
X_t = (\Phi_t,\,\mathbf{v}_t,\,S_t)
\quad\text{(scalar, vector, entropy)}
\]

\[
\Omega_t = \text{admissible future transitions}
\]

\[
H_t = \{e_0,e_1,\dots,e_t\}
\quad\text{(irreversible event log)}
\]

\[
\Psi_t = \text{preference / constraint field}
\]

\[
\M_t : \Omega_t \to \mathbb{R}
\quad\text{(valuation measure)}
\]

\[
\boxed{\Omega_t \;\longrightarrow\; H_{t+1}}
\quad\text{(irreversible selection)}
\]

\[
\textbf{Reality}
=
\text{selection of possibility into history}
\]

% ============================================================
\section*{2.\enspace RSVP Field Dynamics}

Scalar plenum density:
\[
\partial_t \Phi
=
D_\Phi \nabla^2\Phi
-
\nabla\cdot(\Phi\mathbf{v})
+
R_\Phi
-
\lambda_\Phi S
\]

Vector flow (Navier--Stokes type):
\[
\partial_t \mathbf{v}
=
-(\mathbf{v}\cdot\nabla)\mathbf{v}
+
\nu\nabla^2\mathbf{v}
-
\nabla\Phi
+
\mathcal{T}(\mathbf{v})
\]

Entropy production--dissipation:
\[
\partial_t S
=
D_S\nabla^2S
+
\sigma(\Phi,\mathbf{v})
-
\gamma S
\]

Coherence conditions:
\[
\mathcal{U}_\omega \subset X_t,
\qquad
\Phi \uparrow,
\quad
\|\mathbf{v}\|_{\mathrm{aligned}}\uparrow,
\quad
S\downarrow
\]

Admissibility field (Barbour extension):
\[
\Adm(X_t)
=
\bigl\{\,\omega\in\Omega_t
\;\big|\;
\kappa(\omega,H_t)\geq\kappa_*
\,\bigr\}
\]

where $\kappa$ is recursive admissibility and $\kappa_*$ is the viability threshold.
Admissibility replaces entropy as the fundamental cosmological selector.

% ============================================================
\section*{3.\enspace Accessibility \& Temporal Structure}

Global model space (all kinematically possible states):
\[
\fM = \bigl\{s_t \mid t\in\mathbb{R}\bigr\}
\]

History-conditioned accessibility slice:
\[
\A_t
=
\bigl\{s\in\fM\mid s
\text{ reachable given }H_t\bigr\}
\]

Temporal asymmetry (monotone shrinkage):
\[
\A_{t+1}^{\mathrm{consistent}}
\subsetneq
\A_t^{\mathrm{counterfactual}}
\]

The superscripts distinguish the \emph{consistent} slice (states compatible with $H_{t+1}$) from the \emph{counterfactual} envelope (states that were open before the selection at $t$).

Viability (non-collapse):
\[
\Omega_t \neq \emptyset
\]

Provenance-stable self invariant:
\[
\Phi_{\mathrm{self}}
=
\mathrm{Inv}(H_t,\,\Omega_t,\,\A_t,\,\Pp_t)
\]

Shape-space reduction (Barbour):
\[
\fM\big/\!\sim_{\mathrm{gauge}}
\;\longrightarrow\;
\mathcal{Q}_{\mathrm{shape}}
\]

Paths in $\mathcal{Q}_{\mathrm{shape}}$ with maximal $\kappa$
constitute the physical history.

% ============================================================
\section*{4.\enspace Prompting as Field Projection}

Constraint-valid structure space:
$\omega\in\Svalid$

Projection functor:
\[
\Pi:\Svalid\to\mathcal{L}
\]

Rendering operator (stochastic):
\[
\mathcal{R}:\mathcal{L}
\rightsquigarrow
\widehat{\mathcal{L}}
\]

Verification / grounding map:
\[
\mathcal{V}:
\widehat{\mathcal{L}}
\to
\Svalid\cup\{\bot\}
\]

Full pipeline:
\[
\omega
\;\xrightarrow{\;\Pi\;}
p
\;\overset{\mathcal{R}}{\rightsquigarrow}\;
\hat{\ell}
\;\xrightarrow{\;\mathcal{V}\;}
\omega'
\]

Compression chain:
\[
\text{Constraint}
\to
\text{Projection}
\to
\text{Rendering}
\to
\text{Verification}
\]

% ============================================================
\section*{5.\enspace Provenance \& Identity}

Provenance trace:
\[
\Pp_t = \Pp(\omega_t),
\qquad
\Pp_t
\Rightarrow
\omega_t
\]

Identity as maximal integral curve:
\[
\forall t,\quad
\Pp_t
\text{ reconstructs }
\omega_t
\]

Log sovereignty (state determined by history):
\[
X_t = f(H_t)
\]

Append-only event accumulation:
\[
H_{t+1}
=
H_t\cup\{e_t\}
\]

Irreversibility (no right inverse):
\[
\Omega_t\to H_{t+1}
\;\not\Rightarrow\;
H_t
\]

Upload incoherence: no isomorphism $\iota: X_t \xrightarrow{\;\sim\;} X_t'$
preserves $H_t$ under substrate change; identity requires the integral curve, not merely the snapshot.

% ============================================================
\section*{6.\enspace Expansion Operators}

Expansion:
\[
T:\Omega_t\to\Omega_{t+1},
\qquad
T(\Omega_t)
\subseteq
\Omega(H_{t+1})
\]

Operator decomposition:
\[
T
=
T_{\mathrm{phys}}
\circ
T_{\mathrm{comp}}
\circ
T_{\mathrm{sym}}
\circ
T_{\mathrm{sem}}
\]

Spectral transform compatibility:
\[
\mathcal{F}(\Omega_{t+1})
=
\mathcal{T}
\bigl(\mathcal{F}(\Omega_t)\bigr)
\]

Viability under composition:
\[
\Omega_t\neq\emptyset
\;\Rightarrow\;
T(\Omega_t)\neq\emptyset
\]

\[
\boxed{
\textbf{Intelligence}
=
\text{selection of viable expansion operators}
}
\]

% ============================================================
\section*{7.\enspace Sheaf \& Cohomology Structure}

Let $\mathfrak{U}=\{U_i\}$ be an open cover of $\fM$.
Local accessibility sheaf $\mathscr{A}$:
\[
\mathscr{A}(U_i) = \A_t\cap U_i
\]

Gluing condition (consistency of local histories):
\[
\mathscr{A}(U_i)\big|_{U_i\cap U_j}
=
\mathscr{A}(U_j)\big|_{U_i\cap U_j}
\]

Obstruction to global section:
\[
[\,\delta\sigma\,]
\;\in\;
\Hcech^1(\mathfrak{U},\,\mathscr{A})
\]

Vanishing obstruction $\Leftrightarrow$ globally consistent history.
Non-trivial $\Hcech^1$ encodes irreducible temporal contradictions or
constraint-incoherent transitions.

\v{C}ech 2-cocycle (triple overlap):
\[
\delta^2\sigma_{ijk}
=
\sigma_{jk}
-\sigma_{ik}
+\sigma_{ij}
=0
\]

Holonomy of constraint transport around closed loop $\gamma$:
\[
\mathrm{Hol}_\gamma(\Psi)
\;=\;
\mathcal{P}\exp\!\Bigl(
\oint_\gamma A_\Psi\,d\ell
\Bigr)
\;\in\;
\mathrm{Aut}(\Omega_t)
\]

Non-trivial holonomy $\Rightarrow$ axial charge; cf.\ cosmic birefringence as
global holonomy of the admissibility connection over $\mathcal{Q}_{\mathrm{shape}}$.

% ============================================================
\section*{8.\enspace Spherepop / KES Connection}

Spherepop operators over event histories:

\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
Operator & KES interpretation\\
\midrule
$\mathrm{Pop}(e)$ & $H_t\to H_t\cup\{e\}$ (append event)\\
$\mathrm{Bind}(e,c)$ & $\omega\mapsto\omega'$ with constraint $c$\\
$\mathrm{Collapse}(\omega)$ & $\Omega_t\to H_{t+1}$ (actuation)\\
$\mathrm{Refuse}(\omega)$ & $\omega\notin\Adm(X_t)$\\
\bottomrule
\end{tabular}
\end{center}

Free traced monoidal structure:
\[
(\mathbf{Sp},\,\otimes,\,I,\,\mathrm{Tr})
\]

Trace encodes feedback loops; append-only $H_t$ prevents trace from
reversing events. The Spherepop category embeds into $\C$ via:
\[
\iota:\mathbf{Sp}\hookrightarrow\C,
\qquad
\iota(\mathrm{Pop})\mapsto [\,H_t\to H_{t+1}\,]
\]

% ============================================================
\section*{9.\enspace Economic Projection}

Valuation measure:
\[
\M_t:\Omega_t^{\mathrm{agg}}
\to
\mathbb{R}
\]

Lossy compression (distinct states collapse):
\[
\M_t(\omega_1)
=
\M_t(\omega_2),
\quad
\omega_1\neq\omega_2
\]

Extraction regime (Goodhart divergence):
\[
\Delta\M_t >0,
\qquad
\Delta\Rr_t<0
\]

Care functional:
\[
\mathcal{C}_t:
\KES_t\to\mathbb{R},
\qquad
\M_t \uparrow
\;\Rightarrow\;
\mathcal{C}_t \downarrow
\]

Structural richness:
\[
\Rr(\Omega)
=
\dim(\Omega)
+
\mathrm{conn}(\Omega)
+
\mathrm{coh}(\Omega)
\]

Platform extraction reads as $\M_t$ maximisation with
$\Rr(\Omega_{\mathrm{worker}})$ as externality; austerity shrinks
$\A_t^{\mathrm{public}}$ while preserving $\A_t^{\mathrm{capital}}$.

% ============================================================
\section*{10.\enspace Markov Failure Case}

Standard generator:
\[
P(x_{n+1}\mid x_n),
\quad
H_t = \varnothing,
\quad
\Pp_t = \varnothing,
\quad
\mathcal{V} = \varnothing
\]

Consequence:
\[
\hat{\ell}
\;\sim\;
\text{prior manifold}
\]

Surface novelty vs.\ structural distinction:
\[
\delta_{\mathrm{surface}}(\hat{\ell}_1,\hat{\ell}_2) > 0
\;\not\Rightarrow\;
\delta_{\mathrm{structural}}(\omega_1,\omega_2) > 0
\]

Collapse regime: without $H_t$, outputs sample the attractor basin of the
training distribution; constraint-valid generation requires $\mathcal{V}\neq\varnothing$.

% ============================================================
\section*{11.\enspace Spectral Text Structure}

Token embedding sequence:
\[
x(n)
\in
\mathbb{R}^d,
\quad
n = 0,\dots,N-1
\]

Discrete Fourier transform:
\[
X(k)
=
\sum_{n=0}^{N-1}
x(n)\,e^{-i2\pi kn/N}
\]

Spectral signature:
\[
\sigma_{\mathrm{text}}
=
\bigl|X(k)\bigr|
\]

Constraint-preserving transform:
\[
f:\omega\to\omega',
\quad
\mathcal{C}(\omega)=1
\;\Rightarrow\;
\mathcal{C}(\omega')=1
\]

Low-frequency modes carry semantic coherence;
high-frequency modes carry syntactic variation.
Structural distinction lives in $\sigma_{\mathrm{text}}$, not surface tokens.

% ============================================================
\section*{12.\enspace Category Structure}

Possibility category $\C$:
\[
\mathrm{Ob}(\C)=\{\Omega\},
\qquad
\mathrm{Hom}(\Omega,\Omega')=\{\omega:\Omega\to\Omega'\}
\]

History as composition:
\[
H_t
=
\omega_t\circ\cdots\circ\omega_1
\]

Functors:
\[
\Pi:\mathcal{S}\to\mathcal{L}
\quad\text{(projection)},
\qquad
\M_t:\C\to\mathbf{R}
\quad\text{(valuation)}
\]

Natural transformation (power shift):
\[
\eta:\M_t
\Rightarrow
\M_t'
\]

Alignment functor chain:
\[
\mathbf{Val}
\xrightarrow{F}
\mathbf{Beh}
\xrightarrow{G}
\mathbf{Act},
\quad
G\circ F \simeq \mathrm{id}_{\mathbf{Val}}
\quad\text{(alignment condition)}
\]

The weak equivalence $\simeq$ rather than strict equality acknowledges that
behavioural implementation is at best naturally isomorphic to value specification;
strict identity would require a fully transparent action--value correspondence
unavailable under bounded rationality.

% ============================================================
\section*{13.\enspace Master Diagram}

The following square unifies the six primary KES relations---possibility,
selection, provenance, projection, rendering, and verification---in a single
commuting structure. The bent arrow $\mathcal{V}$ closes the loop from rendered
output back into the constraint-valid possibility space, making the cycle
explicit.

\[
\begin{tikzcd}[column sep=huge, row sep=large]
\Omega_t
\arrow[r,"\mathrm{selection}"]
\arrow[d,"\Pi"']
&
H_{t+1}
\arrow[d,"\mathcal{P}"]
\\
\mathcal{L}
\arrow[r,squiggly,"\mathcal{R}"']
&
\widehat{\mathcal{L}}
\arrow[ul,bend left=20,"\mathcal{V}"]
\end{tikzcd}
\]

Reading the diagram: admissible possibilities $\Omega_t$ are selected into
history $H_{t+1}$ (top); simultaneously projected via $\Pi$ into linguistic
space $\mathcal{L}$ (left); stochastically rendered to $\widehat{\mathcal{L}}$
(bottom); and verified by $\mathcal{V}$ back into the next possibility set
(diagonal). Provenance $\mathcal{P}$ (right) records the selection event,
coupling $H_{t+1}$ to the rendered output and grounding the loop.

% ============================================================
\section*{14.\enspace Fundamental Diagrams}

Projection--rendering pipeline:
\[
\begin{tikzcd}[column sep=1.6em]
\omega
\arrow[r,"\Pi"]
&
p
\arrow[r,"\mathcal{R}",squiggly]
&
\hat{\ell}
\arrow[r,"\mathcal{V}"]
&
\omega'
\end{tikzcd}
\]

Temporal shrinkage:
\[
\begin{tikzcd}[column sep=1.6em]
\Omega_t
\arrow[r]
&
H_{t+1}
\arrow[r]
&
\A_{t+1}^{\mathrm{consistent}}
\;\subset\;
\A_t^{\mathrm{counterfactual}}
\end{tikzcd}
\]

Full rendering cycle:
\[
\begin{tikzcd}[column sep=1.4em]
\Svalid
\arrow[r,"\Pi"]
&
\mathcal{L}
\arrow[r,squiggly,"\mathcal{R}"]
&
\widehat{\mathcal{L}}
\arrow[r,"\mathcal{V}"]
&
\Svalid
\cup\{\bot\}
\end{tikzcd}
\]

Sheaf gluing pullback:
\[
\begin{tikzcd}
\mathscr{A}(U_i\cup U_j)
\arrow[r]
\arrow[d]
&
\mathscr{A}(U_i)
\arrow[d]
\\
\mathscr{A}(U_j)
\arrow[r]
&
\mathscr{A}(U_i\cap U_j)
\end{tikzcd}
\]

% ============================================================

\end{multicols}

\vspace{0.3em}
\noindent\rule{\linewidth}{0.4pt}
\vspace{0.1em}

{\centering\small\textbf{15.\enspace Compression Statements}\par}
\vspace{0.3em}

\begin{mdframed}[style=compbox]
\setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{2pt}
\setlength{\abovedisplayshortskip}{1pt}
\setlength{\belowdisplayshortskip}{1pt}
\begin{align*}
\textbf{Time}        &= \text{constraint accumulation} \quad&
\textbf{Mind}        &= \text{local accessibility with provenance} \\
\textbf{Self}        &= \text{provenance-stable integral curve} \quad&
\textbf{Intelligence}&= \text{coherence-preserving expansion} \\
\textbf{Prompting}   &= \text{field-guided projection} \quad&
\textbf{Economics}   &= \text{compressed possibility valuation} \\
\textbf{Care}        &= \text{preservation of structural richness} \quad&
\textbf{Reality}     &= \Omega_t\;\longrightarrow\;H_{t+1} \\
\textbf{Identity}    &= \text{maximal integral curve through }H_t \quad&
\textbf{Cosmology}   &= \text{admissibility geometry on }\mathcal{Q}_{\mathrm{shape}} \\
\textbf{Power}       &= \eta:\M_t\Rightarrow\M_t' \quad&
\textbf{Collapse}    &= \rho(D\mathcal{G}_{\KES^*}) > 1
\end{align*}
\end{mdframed}

\vspace{0.5em}

\begin{multicols}{2}

% ============================================================
\section*{16.\enspace Recursive Constraint Geometry}

Constraint propagation over admissible trajectories:
\[
\mathcal{C}_{t+1}(\omega)
=
\mathcal{C}_t(\omega)
+
\Delta_{\mathrm{hist}}(\omega,H_t)
-
\Delta_{\mathrm{entropy}}(\omega)
\]

Recursive admissibility flow:
\[
\kappa_{t+1}(\omega)
=
\kappa_t(\omega)
+
\int_{\gamma_\omega}
\Psi_t \cdot d\ell
-
S_t(\omega)
\]

Constraint curvature tensor:
\[
\mathfrak{R}_{ij}
=
\partial_i\Psi_j
-
\partial_j\Psi_i
+
[\Psi_i,\Psi_j]
\]

Flat constraint regions satisfy $\mathfrak{R}_{ij}=0$;
nonzero curvature yields path-ordering dependence:
\[
\mathfrak{R}_{ij}\neq0
\;\Rightarrow\;
\omega_a\circ\omega_b
\neq
\omega_b\circ\omega_a
\]

\begin{corollary}
Historical ordering is dynamically significant precisely when
constraint curvature is non-vanishing.
\end{corollary}

\begin{corollary}
Regions of high recursive curvature correspond to strongly irreversible
civilizational or cognitive transitions.
\end{corollary}

% ============================================================
\section*{17.\enspace Provenance Entropy \& Information Flow}

Provenance entropy:
\[
S_{\Pp}(H_t)
=
-\sum_i
p_i\log p_i
\]

where $p_i$ is the recoverability weight of provenance branch $i$.

Recoverability functional:
\[
\mathcal{Q}_{\mathrm{rec}}
=
\frac{
\dim(\mathrm{Recoverable}(H_t))
}{
\dim(H_t)
}
\]

Epistemic fragmentation condition:
\[
\partial_t\mathcal{Q}_{\mathrm{rec}} < 0
\]

Information transport equation:
\[
\partial_t I
+
\nabla\cdot(I\mathbf{v})
=
D_I\nabla^2 I
-
\lambda_I S
\]

Constraint-preserving channels satisfy:
\[
\frac{d}{dt}\mathcal{C}(\omega_t)\geq0
\]

\begin{corollary}
Compression systems maximising transmission while minimising provenance
necessarily increase semantic ambiguity.
\end{corollary}

\begin{corollary}
A civilisation with declining provenance recoverability will increasingly
substitute symbolic authority for reconstructible derivation.
\end{corollary}

% ============================================================
\section*{18.\enspace Narrative Type Theory}

Let $\mathbb{T}$ be the category of admissible narrative types.
Objects $\tau\in\mathrm{Ob}(\mathbb{T})$ are stable narrative invariants;
morphisms $f:\tau_i\to\tau_j$ are admissible identity transitions.

Narrative typing judgment:
\[
H_t \vdash \tau
\]

Consistency condition:
\[
H_t \vdash \tau_i
\;\wedge\;
\tau_i\not\leadsto\tau_j
\;\Rightarrow\;
H_t \nvdash \tau_j
\]

Narrative bifurcation operator:
\[
\mathcal{B}:\tau\mapsto\{\tau_1,\tau_2,\dots\}
\]

Self-coherence invariant:
\[
\chi_{\mathrm{self}}
=
\int_{H_t}
\mathrm{Compat}(\tau_t,\tau_{t+1})\,dt
\]

\begin{corollary}
Psychological fragmentation corresponds to decreasing path-compatibility
within the narrative type manifold.
\end{corollary}

\begin{corollary}
Stable identity is not static persistence but coherent type evolution
through irreversible history.
\end{corollary}

% ============================================================
\section*{19.\enspace Spectral Constraint Topology}

Decompose $\Omega_t=\bigcup_i\Omega_i$ into local admissibility domains.

Spectral Laplacian:
\[
\Delta_\Omega = D - A
\]

where $D$ is the degree operator and $A$ the accessibility adjacency.

Eigenstructure:
\[
\Delta_\Omega\phi_n = \lambda_n\phi_n
\]

Low-frequency modes ($\lambda_n\approx0$) encode globally coherent
semantic structures; high-frequency modes ($\lambda_n\gg0$) encode
localised fluctuations or instability.

Constraint resonance condition:
\[
\langle\phi_i,\Psi_t\rangle \gg 0
\]

Phase-locking:
\[
\phi_i(t+\Delta t)
\approx
e^{i\theta}\phi_i(t)
\]

\begin{corollary}
Long-term civilisations stabilise around low-frequency admissibility
eigenmodes.
\end{corollary}

\begin{corollary}
Memetic instability corresponds to excitation of high-frequency spectral
components exceeding damping capacity.
\end{corollary}

% ============================================================
\section*{20.\enspace The KES Fixed-Point Theorem}

Recursive evolution operator:
\[
\KES_{t+1} = \mathcal{G}(\KES_t),
\qquad
\KES^* = \mathcal{G}(\KES^*)
\]

Local stability via spectral radius $\rho$:

\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
Condition & Regime \\
\midrule
$\rho(D\mathcal{G}_{\KES^*})<1$ & Stable expansion \\
$\rho\approx1$ & Metastable \\
$\rho>1$ & Collapse \\
\bottomrule
\end{tabular}
\end{center}

\begin{theorem}[Recursive Coherence Theorem]
A KES trajectory remains structurally viable iff recursive admissibility,
provenance recoverability, and expansion coherence remain jointly bounded.
\end{theorem}

\begin{proof}[Sketch]
If admissibility collapses, the trajectory exits $\Adm(X_t)$.
If provenance collapses, reconstruction of $\omega_t$ from $\Pp_t$ fails.
If expansion coherence collapses, $\Omega_t$ fragments into disconnected
components. Joint boundedness is therefore necessary for persistent
structural identity.
\end{proof}

\begin{corollary}
Intelligence is not maximal optimisation but bounded coherent continuation
through recursively constrained possibility space.
\end{corollary}

\begin{corollary}
The terminal failure mode of civilisations is topological fragmentation of
their admissibility manifold.
\end{corollary}

\end{multicols}

\end{document}
